---
title: "DS: Project 2"
output: html_notebook
---

```{r basic, include=F}
# use this function to conveniently load libraries and work smoothly with knitting
# can add quietly=T option to the require() function
# the loadPkg function essentially replaced/substituted two functions install.packages() and library() in one step.
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }

# unload/detact package when done using it
unloadPkg = function(pkg, character.only = FALSE) { 
  if(!character.only) { pkg <- as.character(substitute(pkg)) } 
  search_item <- paste("package", pkg,sep = ":") 
  while(search_item %in% search()) { detach(search_item, unload = TRUE, character.only = TRUE) } 
}
```

```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# knitr::opts_chunk$set(include = F)
# knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r xkablesummary}
loadPkg("xtable")
loadPkg("kableExtra")
loadPkg("stringi")

xkabledply = function(modelsmmrytable, title="Table", digits = 4, pos="left", bso="striped", wide=FALSE) { 
  #' Combining base::summary, xtable, and kableExtra, to easily display model summary. 
  #' wrapper for the base::summary function on model objects
  #' Can also use as head for better display
  #' ELo 202004 GWU DATS
  #' version 1.2
  #' @param modelsmmrytable This can be a generic table, a model object such as lm(), or the summary of a model object summary(lm()) 
  #' @param title Title of table. 
  #' @param digits Number of digits to display
  #' @param pos Position of table, c("left","center","right") 
  #' @param bso bootstrap_options = c("basic", "striped", "bordered", "hover", "condensed", "responsive")
  #' @param wide print table in long (FALSE) format or wide (TRUE) format
  #' @return HTML table for display
  #' @examples
  #' library("xtable")
  #' library("kableExtra")
  #' xkabledply( df, title="Table testing", pos="left", bso="hover" )
  #' xkabledply( ISLR::Hitters[1:5,] )
  if (wide) { modelsmmrytable <- t(modelsmmrytable) }
  modelsmmrytable %>%
    xtable() %>% 
    kable(caption = title, digits = digits) %>%
    kable_styling(bootstrap_options = bso, full_width = FALSE, position = pos)
}

xkabledplyhead = function(df, rows=5, title="Head", digits = 4, pos="left", bso="striped") { 
  xkabledply(df[1:rows, ], title, digits, pos, bso, wide=FALSE)
}

xkabledplytail = function(df, rows=5, title="Tail", digits = 4, pos="left", bso="striped") { 
  trows = nrow(df)
  xkabledply(df[ (trows-rows+1) : trows, ], title, digits, pos, bso, wide=FALSE)
}

xkablesummary = function(df, title="Table: Statistics summary.", digits = 4, pos="left", bso="striped") { 
  #' Combining base::summary, xtable, and kableExtra, to easily display numeric variable summary of dataframes. 
  #' ELo 202004 GWU DATS
  #' version 1.2
  #' @param df The dataframe.
  #' @param title Title of table. 
  #' @param digits Number of digits to display
  #' @param pos Position of table, c("left","center","right") 
  #' @param bso bootstrap_options = c("basic", "striped", "bordered", "hover", "condensed", "responsive")
  #' @return The HTML summary table for display, or for knitr to process into other formats 
  #' @examples
  #' xkablesummary( faraway::ozone )
  #' xkablesummary( ISLR::Hitters, title="Five number summary", pos="left", bso="hover"  )
  
  s = summary(df) %>%
    apply( 2, function(x) stringr::str_remove_all(x,c("Min.\\s*:\\s*","1st Qu.\\s*:\\s*","Median\\s*:\\s*","Mean\\s*:\\s*","3rd Qu.\\s*:\\s*","Max.\\s*:\\s*")) ) %>% # replace all leading words
    apply( 2, function(x) stringr::str_trim(x, "right")) # trim trailing spaces left
  
  colnames(s) <- stringr::str_trim(colnames(s))
  
  if ( dim(s)[1] ==6 ) { rownames(s) <- c('Min','Q1','Median','Mean','Q3','Max') 
  } else if ( dim(s)[1] ==7 ) { rownames(s) <- c('Min','Q1','Median','Mean','Q3','Max','NA') }
  
  xkabledply(s, title=title, digits = digits, pos=pos, bso=bso )
}

xkablevif = function(model, title="VIFs of the model", digits = 3, pos="left", bso="striped", wide=TRUE) { 
  #' Combining faraway::vif, xtable, and kableExtra, to easily display numeric summary of VIFs for a model. 
  #' ELo 202004 GWU DATS
  #' version 1.2
  #' @param model The lm or compatible model object.
  #' @param title Title of table. 
  #' @param digits Number of digits to display
  #' @param pos Position of table, c("left","center","right") 
  #' @param bso bootstrap_options = c("basic", "striped", "bordered", "hover", "condensed", "responsive")
  #' @param wide print table in long (FALSE) format or wide (TRUE) format
  #' @return The HTML summary table of the VIFs for a model for display, or for knitr to process into other formats 
  #' @examples
  #' xkablevif( lm(Salary~Hits+RBI, data=ISLR::Hitters), wide=T )
  
  vifs = table( names(model$coefficients)[2:length(model$coefficients)] ) # remove intercept to set column names
  vifs[] = faraway::vif(model) # set the values
  if (wide) { vifs <- t(vifs) }
  xkabledply( vifs, title=title, digits = digits, pos=pos, bso=bso )
}
```



Our Data set consists of 54 variables with 3941 observations.
Upon closer inspection we can see that there are 17 variables related to Alcohol consuption, 16 variables regarding Cannabis use, 10 demographic descriptive variables, and 9 variables related to the individuals' psychological aptitudes during the covid-19 pandemic. 

```{r}
library(readr)
Substance_Use <- read_csv("Substance Use.csv")
head(Substance_Use)
str(Substance_Use)

Alcohol <- subset(Substance_Use[10:26])
Cannabis <- subset(Substance_Use[27:42])
Pandemic_Attitudes <- subset(Substance_Use[2:9])
Demographics <- subset(Substance_Use[43:50])
```


Exploring demographic categories and mental health during the pandemic

```{r}
#Employment and pandemic
cor.test(Substance_Use$PEMPSTC, Substance_Use$ME_10)
cor.test(Substance_Use$PEMPSTC, Substance_Use$ME_15)

#age and pandemic 
cor.test(Substance_Use$AGEGRP, Substance_Use$ME_10)
cor.test(Substance_Use$AGEGRP, Substance_Use$ME_15)

#Marital status and pandemic
cor.test(Substance_Use$PMARSTC, Substance_Use$ME_10)
cor.test(Substance_Use$PMARSTC, Substance_Use$ME_15)

#household size and pandemic
cor.test(Substance_Use$PHHLDSZC, Substance_Use$ME_10)
cor.test(Substance_Use$PHHLDSZC, Substance_Use$ME_15)
```

Alcohol consumption and demographic characteristics:

```{r}
Alcohol_correlations <- cor(Alcohol)
xkabledply(Alcohol_correlations)

Alcohol_Pandemic <- merge(Alcohol, Pandemic_Attitudes)

Alcohol_Pandemic_Cor <- cor(Alcohol_Pandemic)
xkabledply(Alcohol_Pandemic_Cor)

```

```{r}
Alc_Age <- lm(ALC_10C ~ AGEGRP, data = Substance_Use)
summary(Alc_Age)
```

```{r}
Alc_Marriage <- lm(ALC_10C ~ PMARSTC, data = Substance_Use)
summary(Alc_Marriage)
```

```{r}
Alc_Education <- lm(ALC_10C ~ PEDUC_LC, data = Substance_Use)
summary(Alc_Education)
```

```{r}
Alc_Employment <- lm(ALC_10C ~ PEMPSTC, data = Substance_Use)
summary(Alc_Employment)
```

```{r}
Alc_Sex <- lm(ALC_10C ~ SEX, data = Substance_Use)
summary(Alc_Sex)
```

```{r}
Alc_Setting <- lm(ALC_10C ~ RURURB, data = Substance_Use)
summary(Alc_Setting)
```

Alcohol and Mental Health during the pandemic:

```{r}
Increased_ALC_MHealth <- lm(ME_10 ~ ALC_10C, data = Alcohol_Pandemic)
summary(Increased_ALC_MHealth)
```

```{r}
Increased_ALC_Satisfaction <- lm(LS_05 ~ ALC_10C, data = Alcohol_Pandemic)
summary(Increased_ALC_Satisfaction)
```

```{r}
Increased_ALC_Stress <- lm(ME_15 ~ ALC_10C, data = Alcohol_Pandemic)
summary(Increased_ALC_Stress)
```

```{r}
Increased_ALC_SocialSupport <- lm(ME_20 ~ ALC_10C, data = Alcohol_Pandemic)
summary(Increased_ALC_SocialSupport)
```

```{r}
Increased_ALC_Loneliness <- lm(ALC_10C ~ ME_30A + ME_30C, data = Alcohol_Pandemic)
summary(Increased_ALC_Loneliness)
```


Cannabis and consumption and demographic characteristics:

```{r}
Cannabis_correlations <- cor(Cannabis)
xkabledply(Cannabis_correlations)

Cannabis_Pandemic <- merge(Cannabis, Pandemic_Attitudes)

Cannabis_Pandemic_Cor <- cor(Cannabis_Pandemic)
xkabledply(Cannabis_Pandemic_Cor)
```


```{r}
Can_Age <- lm(CAN_15 ~ AGEGRP, data = Substance_Use)
summary(Can_Age)
```

```{r}
Can_Marriage <- lm(CAN_15 ~ PMARSTC, data = Substance_Use)
summary(Can_Marriage)
```

```{r}
Can_Education <- lm(CAN_15 ~ PEDUC_LC, data = Substance_Use)
summary(Can_Education)
```

```{r}
Can_Employment <- lm(CAN_15 ~ PEMPSTC, data = Substance_Use)
summary(Alc_Employment)
```

```{r}
Can_Sex <- lm(CAN_15 ~ SEX, data = Substance_Use)
summary(Can_Sex)
```

```{r}
Can_Setting <- lm(CAN_15 ~ RURURB, data = Substance_Use)
summary(Can_Setting)
```

Cannabis and Mental Health during the pandemic:

```{r}
Increased_CAN_MHealth <- lm(ME_10 ~ CAN_15, data = Cannabis_Pandemic)
summary(Increased_CAN_MHealth)
```

```{r}
Increased_CAN_Satisfaction <- lm(LS_05 ~ CAN_15, data = Cannabis_Pandemic)
summary(Increased_CAN_Satisfaction)
```

```{r}
Increased_CAN_Stress <- lm(ME_15 ~ CAN_15, data = Cannabis_Pandemic)
summary(Increased_CAN_Stress)
```

```{r}
Increased_CAN_SocialSupport <- lm(ME_20 ~ CAN_15, data = Cannabis_Pandemic)
summary(Increased_CAN_SocialSupport)
```

```{r}
Increased_CAN_Loneliness <- lm(CAN_15 ~ ME_30A + ME_30C, data = Cannabis_Pandemic)
summary(Increased_CAN_Loneliness)
```

