---
title: "Introduction to Data Science - Project II"
author: "Team 5: Dreispiel Juan Nicole, Espinosa Robles Manuel, Motiani Sonam, Nieves Pérez Luisa Fernanda"
date: "today"
# date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
    theme: united
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("easypackages")
easypackages::packages("data.table", "DT", "dplyr", "mgcv", "ggplot2", "RColorBrewer", "plotly", "parallel", "mltools", prompt = F)
```


```{r basic, include=F}
# use this function to conveniently load libraries and work smoothly with knitting
# can add quietly=T option to the require() function
# the loadPkg function essentially replaced/substituted two functions install.packages() and library() in one step.
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }

# unload/detact package when done using it
unloadPkg = function(pkg, character.only = FALSE) { 
  if(!character.only) { pkg <- as.character(substitute(pkg)) } 
  search_item <- paste("package", pkg,sep = ":") 
  while(search_item %in% search()) { detach(search_item, unload = TRUE, character.only = TRUE) } 
}
```

```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
# knitr::opts_chunk$set(warning = F, results = "hide", message = F)
# knitr::opts_chunk$set(include = F)
# knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r xkablesummary}
loadPkg("xtable")
loadPkg("kableExtra")
loadPkg("stringi")

xkabledply = function(modelsmmrytable, title="Table", digits = 4, pos="left", bso="striped", wide=FALSE) { 
  #' Combining base::summary, xtable, and kableExtra, to easily display model summary. 
  #' wrapper for the base::summary function on model objects
  #' Can also use as head for better display
  #' ELo 202004 GWU DATS
  #' version 1.2
  #' @param modelsmmrytable This can be a generic table, a model object such as lm(), or the summary of a model object summary(lm()) 
  #' @param title Title of table. 
  #' @param digits Number of digits to display
  #' @param pos Position of table, c("left","center","right") 
  #' @param bso bootstrap_options = c("basic", "striped", "bordered", "hover", "condensed", "responsive")
  #' @param wide print table in long (FALSE) format or wide (TRUE) format
  #' @return HTML table for display
  #' @examples
  #' library("xtable")
  #' library("kableExtra")
  #' xkabledply( df, title="Table testing", pos="left", bso="hover" )
  #' xkabledply( ISLR::Hitters[1:5,] )
  if (wide) { modelsmmrytable <- t(modelsmmrytable) }
  modelsmmrytable %>%
    xtable() %>% 
    kable(caption = title, digits = digits) %>%
    kable_styling(bootstrap_options = bso, full_width = FALSE, position = pos)
}

xkabledplyhead = function(df, rows=5, title="Head", digits = 4, pos="left", bso="striped") { 
  xkabledply(df[1:rows, ], title, digits, pos, bso, wide=FALSE)
}

xkabledplytail = function(df, rows=5, title="Tail", digits = 4, pos="left", bso="striped") { 
  trows = nrow(df)
  xkabledply(df[ (trows-rows+1) : trows, ], title, digits, pos, bso, wide=FALSE)
}

xkablesummary = function(df, title="Table: Statistics summary.", digits = 4, pos="left", bso="striped") { 
  #' Combining base::summary, xtable, and kableExtra, to easily display numeric variable summary of dataframes. 
  #' ELo 202004 GWU DATS
  #' version 1.2
  #' @param df The dataframe.
  #' @param title Title of table. 
  #' @param digits Number of digits to display
  #' @param pos Position of table, c("left","center","right") 
  #' @param bso bootstrap_options = c("basic", "striped", "bordered", "hover", "condensed", "responsive")
  #' @return The HTML summary table for display, or for knitr to process into other formats 
  #' @examples
  #' xkablesummary( faraway::ozone )
  #' xkablesummary( ISLR::Hitters, title="Five number summary", pos="left", bso="hover"  )
  
  s = summary(df) %>%
    apply( 2, function(x) stringr::str_remove_all(x,c("Min.\\s*:\\s*","1st Qu.\\s*:\\s*","Median\\s*:\\s*","Mean\\s*:\\s*","3rd Qu.\\s*:\\s*","Max.\\s*:\\s*")) ) %>% # replace all leading words
    apply( 2, function(x) stringr::str_trim(x, "right")) # trim trailing spaces left
  
  colnames(s) <- stringr::str_trim(colnames(s))
  
  if ( dim(s)[1] ==6 ) { rownames(s) <- c('Min','Q1','Median','Mean','Q3','Max') 
  } else if ( dim(s)[1] ==7 ) { rownames(s) <- c('Min','Q1','Median','Mean','Q3','Max','NA') }
  
  xkabledply(s, title=title, digits = digits, pos=pos, bso=bso )
}

xkablevif = function(model, title="VIFs of the model", digits = 3, pos="left", bso="striped", wide=TRUE) { 
  #' Combining faraway::vif, xtable, and kableExtra, to easily display numeric summary of VIFs for a model. 
  #' ELo 202004 GWU DATS
  #' version 1.2
  #' @param model The lm or compatible model object.
  #' @param title Title of table. 
  #' @param digits Number of digits to display
  #' @param pos Position of table, c("left","center","right") 
  #' @param bso bootstrap_options = c("basic", "striped", "bordered", "hover", "condensed", "responsive")
  #' @param wide print table in long (FALSE) format or wide (TRUE) format
  #' @return The HTML summary table of the VIFs for a model for display, or for knitr to process into other formats 
  #' @examples
  #' xkablevif( lm(Salary~Hits+RBI, data=ISLR::Hitters, wide=T ) )
  
  vifs = table( names(model$coefficients)[2:length(model$coefficients)] ) # remove intercept to set column names
  vifs[] = faraway::vif(model) # set the values
  if (wide) { vifs <- t(vifs) }
  xkabledply( vifs, title=title, digits = digits, pos=pos, bso=bso )
}
```

```{r outlierKD2}
# Fix outliers
outlierKD2 <- function(df, var, rm=FALSE, boxplt=FALSE, histogram=TRUE, qqplt=FALSE) { 
    #' Original outlierKD functino by By Klodian Dhana,
    #' https://www.r-bloggers.com/identify-describe-plot-and-remove-the-outliers-from-the-dataset/
    #' Modified to have third argument for removing outliers instead of interactive prompt, 
    #' and after removing outlier, original df will not be changed. The function returns the a df, 
    #' which can be saved as original df name if desired.
    #' Also added QQ-plot in the output, with options to show/hide boxplot, histogram, qqplot.
    #' Check outliers, and option to remove them, save as a new dataframe. 
    #' @param df The dataframe.
    #' @param var The variable in the dataframe to be checked for outliers
    #' @param rm Boolean. Whether to remove outliers or not.
    #' @param boxplt Boolean. Whether to show the boxplot, before and after outliers removed.
    #' @param histogram Boolean. Whether to show the histogram, before and after outliers removed.
    #' @param qqplt Boolean. Whether to show the qqplot, before and after outliers removed.
    #' @return The dataframe with outliers replaced by NA if rm==TRUE, or df if nothing changed
    #' @examples
    #' outlierKD2(mydf, height, FALSE, TRUE, TRUE, TRUE)
    #' mydf = outlierKD2(mydf, height, TRUE, TRUE, TRUE, TRUE)
    #' mydfnew = outlierKD2(mydf, height, TRUE)
    dt = df # duplicate the dataframe for potential alteration
    var_name <- eval(substitute(var),eval(dt))
    na1 <- sum(is.na(var_name))
    m1 <- mean(var_name, na.rm = T)
    par(mfrow=c(2, boxplt+histogram+qqplt), oma=c(0,0,3,0))
    if (qqplt) { 
      qqnorm(var_name, main = "With outliers")
      qqline(var_name)
    }
    if (histogram) { hist(var_name, main="With outliers", xlab=NA, ylab=NA) }
    if (boxplt) { boxplot(var_name, main="With outliers") }

    outlier <- boxplot.stats(var_name)$out
    mo <- mean(outlier)
    var_name <- ifelse(var_name %in% outlier, NA, var_name)
    if (qqplt) { 
      qqnorm(var_name, main = "Without outliers")
      qqline(var_name)
    }
    if (histogram) { hist(var_name, main="Without outliers", xlab=NA, ylab=NA) }
    if (boxplt) { boxplot(var_name, main="Without outliers") }

    title("Outlier Check", outer=TRUE)
    na2 <- sum(is.na(var_name))
    cat("Outliers identified:", na2 - na1, "\n")
    cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "\n")
    cat("Mean of the outliers:", round(mo, 2), "\n")
    m2 <- mean(var_name, na.rm = T)
    cat("Mean without removing outliers:", round(m1, 2), "\n")
    cat("Mean if we remove outliers:", round(m2, 2), "\n")
    
    # response <- readline(prompt="Do you want to remove outliers and to replace with NA? [yes/no]: ")
    # if(response == "y" | response == "yes"){
    if(rm){
        dt[as.character(substitute(var))] <- invisible(var_name)
        #assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
        cat("Outliers successfully removed", "\n")
        return(invisible(dt))
    } else {
        cat("Nothing changed", "\n")
        return(invisible(df))
    }
}
```


###Introduction

Our Data set consists of 54 variables with 3941 observations.
Upon closer inspection we can see that there are 17 variables related to Alcohol consuption, 16 variables regarding Cannabis use, 10 demographic descriptive variables, and 9 variables related to the individuals' psychological aptitudes during the covid-19 pandemic. 

```{r}
library(readr)
Substance_Use <- read_csv("Substance Use.csv")
head(Substance_Use)
#str(Substance_Use)

#stats_with function is so that we don't have to indvidually write all variables that start with prefix "CAN" or "ALC"
main_clean <- Substance_Use %>% select (AGEGRP, PHHLDSZC, IMMIGRNC, SEX, PMARSTC, PEMPSTC, PEDUC_LC, RURURB, STRESS, LS_05,  starts_with('ME'), starts_with('CAN'), starts_with('ALC'), PERS_WGT)

#head(main_clean)
#str(main_clean)

Alcohol <- subset(Substance_Use[10:26])
Cannabis <- subset(Substance_Use[27:42])
Pandemic_Attitudes <- subset(Substance_Use[2:9])
Demographics <- subset(Substance_Use[43:50])
```

###EDA

***Exploring demographic categories and mental health during the pandemic***
We performed a series of correlations between the variables and throughout categories to expose the significant values and their directional influence (positive or negative). 
We found the following to be the most highly correlated variables between mental health and demographics were:
•	Mental health(ME_10)
o	AGEGRP: -0.101
•	Stress(ME_15)
o	AGEGRP: -0.223

```{r}
#Employment and pandemic
cor.test(Substance_Use$PEMPSTC, Substance_Use$ME_10)
cor.test(Substance_Use$PEMPSTC, Substance_Use$ME_15)

#age and pandemic 
cor.test(Substance_Use$AGEGRP, Substance_Use$ME_10)
cor.test(Substance_Use$AGEGRP, Substance_Use$ME_15)

#Marital status and pandemic
cor.test(Substance_Use$PMARSTC, Substance_Use$ME_10)
cor.test(Substance_Use$PMARSTC, Substance_Use$ME_15)

#household size and pandemic
cor.test(Substance_Use$PHHLDSZC, Substance_Use$ME_10)
cor.test(Substance_Use$PHHLDSZC, Substance_Use$ME_15)

#Education Level and pandemic
cor.test(Substance_Use$PEDUC_LC, Substance_Use$ME_10)
cor.test(Substance_Use$PEDUC_LC, Substance_Use$ME_15)

#Sex and pandemic
cor.test(Substance_Use$SEX, Substance_Use$ME_10)
cor.test(Substance_Use$SEX, Substance_Use$ME_15)

#Rural/urban and pandemic
cor.test(Substance_Use$RURURB, Substance_Use$ME_10)
cor.test(Substance_Use$RURURB, Substance_Use$ME_15)

#immigration status and pandemic
cor.test(Substance_Use$IMMIGRNC, Substance_Use$ME_10)
cor.test(Substance_Use$IMMIGRNC, Substance_Use$ME_15)
```


##Correlations

***Alcohol Correlations***
To further understand the impact Covid 19 had on alcohol consumption, we performed a correlation of all the Alcohol variables to explore the posible relationships between the consumption and psychological state of the individuals.
We found that increased alcohol consumption during covid quarantine was negatively correlated with mental health (ME_10). And that increased alcohol consumption also shows a weak positive correlation with immigration (0.075). 
AGEGRP(0.119)
ME_10(-0.133)
Isolation(ME_30C): -0.124
ME_15: -0.118

```{r}
Alcohol_correlations <- cor(Alcohol)
xkabledply(Alcohol_correlations)
```


Alcohol and Demographic Variables
```{r}
cor.test(Alcohol$ALC_20,Demographics$PHHLDSZC )
cor.test(Alcohol$ALC_20, Demographics$AGEGRP)
cor.test(Alcohol$ALC_20, Demographics$IMMIGRNC)
cor.test(Alcohol$ALC_20, Demographics$SEX)
cor.test(Alcohol$ALC_20, Demographics$PMARSTC)
cor.test(Alcohol$ALC_20, Demographics$PEDUC_LC)
cor.test(Alcohol$ALC_20, Demographics$RURURB)
cor.test(Alcohol$ALC_20, Demographics$PEMPSTC)
```

***Alcohol and Mental Health during the pandemic***

```{r}
#Isolation
cor.test(Substance_Use$ALC_20, Substance_Use$ME_30C)

#Mental Health
cor.test(Substance_Use$ALC_20, Substance_Use$ME_10)

#Satisfaction
cor.test(Substance_Use$ALC_20, Substance_Use$LS_05)

#stress
cor.test(Substance_Use$ALC_20, Substance_Use$ME_15)

#Social Support
cor.test(Substance_Use$ALC_20, Substance_Use$ME_20)

```


***Cannabis correlations***
ME_30C (-0.107)
ME_10(-0.101)
ME_15(-0.117)
AGEGRP(0.245)
PMARSTC(-0.138)

```{r}
Cannabis_correlations <- cor(Cannabis)
xkabledply(Cannabis_correlations)
```

***Cannabis and Mental Health during the pandemic***


```{r}
#Isolation
cor.test(Substance_Use$CAN_15, Substance_Use$ME_30C)

#Mental Health
cor.test(Substance_Use$CAN_15, Substance_Use$ME_10)

#Satisfaction
cor.test(Substance_Use$CAN_15, Substance_Use$LS_05)

#stress
cor.test(Substance_Use$CAN_15, Substance_Use$ME_15)

#Social Support
cor.test(Substance_Use$CAN_15, Substance_Use$ME_20)
```

Cannabis and Demographic Variables
```{r}
cor.test(Cannabis$CAN_15,Demographics$PHHLDSZC )
cor.test(Cannabis$CAN_15, Demographics$AGEGRP)
cor.test(Cannabis$CAN_15, Demographics$IMMIGRNC)
cor.test(Cannabis$CAN_15, Demographics$SEX)
cor.test(Cannabis$CAN_15, Demographics$PMARSTC)
cor.test(Cannabis$CAN_15, Demographics$PEDUC_LC)
cor.test(Cannabis$CAN_15, Demographics$RURURB)
cor.test(Cannabis$CAN_15, Demographics$PEMPSTC)
```

###Linear Regression Models###

***Cannabis Model***

Data selection: Demographics and Cannabis

```{r}
#Filter those who never used Canabis, not statued its variable 
#! mean not in 1, 99, so remove those and keep everything else 
cannabis_filtered <- main_clean %>% filter(!CAN_05 %in% c(1,99))

str(cannabis_filtered)

frequency(cannabis_filtered$CAN_05)

frequency(cannabis_filtered$CAN_15)

cannabis_inc <- cannabis_filtered %>% filter(!CAN_15 %in% c(9))
cannabis_inc  <- cannabis_inc %>% mutate(CAN_15_Revised = as.numeric(if_else(CAN_15 == 1,"1","0")), Employed = as.numeric(if_else(PEMPSTC == 4,"0","1")), Marital = as.numeric(if_else(PMARSTC %in% c(1,2), "1", "0")))
cannabis_inc <- cannabis_inc %>% select(-c(CAN_15))

cannabis_dec <- cannabis_filtered %>% filter(!CAN_15 %in% c(9))
cannabis_dec  <- cannabis_dec %>% mutate(CAN_15_Revised = as.numeric(if_else(CAN_15 == 2,"1","0")), Employed = as.numeric(if_else(PEMPSTC == 4,"0","1")))
cannabis_dec <- cannabis_dec %>% select(-c(CAN_15))
```

Model Building
```{r}
library(car)
lmCanDem <- lm(CAN_15_Revised ~ SEX + AGEGRP + RURURB, data = cannabis_inc, weights = PERS_WGT)
summary(lmCanDem)

lmCanMH <- lm(CAN_15_Revised ~ ME_30C + ME_15, data = cannabis_inc)
summary(lmCanMH)

bestlmCan <- lm(CAN_15_Revised ~ AGEGRP + SEX + RURURB + ME_10, data = cannabis_inc) 
anova(bestlmCan)
vif(bestlmCan)

plot(lmCanDem, main = "Linear Regression Cannabis Model", col="turquoise")
```

We revised the CAN_15 variable to assign binary classification whether cannabis consumption has increased or not increased after the pandemic. We then performed a simple linear regression to see which demographic factors are correlated with increased use of cannabis. From the results on a 5% threshold, we can see that sex, age group and rural/urban indicators are the two strongest drivers of increased cannabis use. That is, the younger age groups, females and those residing in urban areas are more likely to increase cannabis consumption after the pandemic. 

We look at the Q-Q plot and the residuals vs leverage plot. Out of approximately 1200 points that can impact the model performance, there are total 3 points above 0.3, one of which is above 0.5. This particular row has more weight than others, as the data assigns different weight to different observations. So this means that this particular observation has a higher weight in the dataset and has the potential to impact the model. 

We also performed a similar simple linear regression on the mental health factors and created a best fit model combining both demographics and mental health variables. 

***ALcohol Model***

Data Selection: Alcohol and Demographics

```{r}
#Filter those who dont drink, not statued its variable 
#! mean not in 3, 99, so remove those and keep everything else 
alcohol_filtered <- main_clean %>% filter(!ALC_05 %in% c(3,99))

str(alcohol_filtered)

frequency(alcohol_filtered$ALC_05)

frequency(alcohol_filtered$ALC_20)


alcohol_inc <- alcohol_filtered %>% filter(!ALC_20 %in% c(9))
alcohol_inc  <- alcohol_inc %>% mutate(ALC_20_Revised = as.numeric(if_else(ALC_20 == 1,"1","0")), Employed = as.numeric(if_else(PEMPSTC == 4,"0","1")), Marital = as.numeric(if_else(PMARSTC %in% c(1,2), "1", "0")))
alcohol_inc <- alcohol_inc %>% select(-c(ALC_20))

alcohol_dec <- alcohol_filtered %>% filter(!ALC_20 %in% c(9))
alcohol_dec <- alcohol_dec %>% mutate(ALC_20_Revised = as.numeric(if_else(ALC_20 == 2,"1","0")), Employed = as.numeric(if_else(PEMPSTC == 4,"0","1")))
alcohol_dec <- alcohol_dec %>% select(-c(ALC_20))
```

Model building
```{r}
lmAlcDem <- lm(ALC_20_Revised ~ SEX + AGEGRP + Marital + Employed + PHHLDSZC + IMMIGRNC + RURURB, data = alcohol_inc, weights = PERS_WGT)
summary(lmAlcDem)

lmAlcMH <- lm(ALC_20_Revised ~ ME_30C + ME_10 + ME_15, data = alcohol_inc)
summary(lmAlcMH)

bestlmAlc <- lm(ALC_20_Revised ~ AGEGRP + PEDUC_LC + ME_30C + ME_10, data = alcohol_inc) 
anova(bestlmAlc)
vif(bestlmAlc)

plot(lmAlcDem, main = "Linear Regression Alcohol Model", col="orange")
```

Just like the variable CAN_15, we revised the variable ALC_20 to assign binary classification whether alcohol consumption has increased or not increased after the pandemic for the respondents. We performed a simple linear regression here as well to see which demographic factors are correlated with increased use of alcohol. From the results on a 5% threshold, we can see that there are several significant drivers of increased alcohol use. These are, sex, age group, household size, marital status, rural/urban settlement, employment and immigration status. Like cannabis, the younger age group and females are more likely to increase alcohol consumption after the pandemic. Likewise, people in smaller household size, living in urban settlement, the ones unemployed, married and born in Canada are more likely to increase alcohol consumption after the pandemic.

Next we created some plots from the model. Looking at the Q-Q plot and the residuals vs leverage plot, here again there are three points that get highlighted (high residual and high leverage). However, there are just these three points and none of these have greater than 0.5 cooks distance thus they are not influential and shouldn’t materially impact our models.

We also performed a similar simple linear regression on the mental health factors and created a best fit model combining both demographics and mental health variables.

###GLM MODELS###

***Alcohol***
```{r}
glmAlcDem <- glm(ALC_20_Revised ~ SEX + AGEGRP + PMARSTC + PEDUC_LC + Employed + PHHLDSZC + IMMIGRNC, data = alcohol_inc, weights = PERS_WGT)
summary(glmAlcDem)
vif(glmAlcDem)

glmAlcMH <- glm(ALC_20_Revised ~ ME_30C + ME_10 + ME_15 + LS_05, data = alcohol_inc, weights = PERS_WGT)
summary(lmAlcMH)
vif(glmAlcMH)
```

Best Alcohol GLM
```{r}
#Select the most significant variables from both demographic and mental health correlations
bestglmAlc <- glm(ALC_20_Revised ~ AGEGRP + SEX + ME_30C + ME_10, data = alcohol_inc)
summary(bestglmAlc)
vif(bestglmAlc)
anova(bestglmAlc)
```

***Cannabis***
```{r}
glmCanDem <- glm(CAN_15_Revised ~ SEX + AGEGRP + PMARSTC + PEDUC_LC + Employed + PHHLDSZC + IMMIGRNC, data = cannabis_inc, weights = PERS_WGT)
summary(glmCanDem)
vif(glmCanDem)

glmCanMH <- glm(CAN_15_Revised ~ ME_30C + ME_10 + ME_15 + LS_05, data = cannabis_inc)
summary(glmCanMH)
vif(glmCanMH)
```

Best Cannabis GLM 
```{r}
#Select the most significant variables from both demographic and mental health correlations
bestglmCan <- glm(CAN_15_Revised ~ AGEGRP + SEX + PHHLDSZC +ME_30C +ME_15, data = cannabis_inc)
summary(bestglmCan)
vif(bestglmCan)
anova(bestglmCan)
```

#PCA Models

Cannabis PCA
```{r}
library(olsrr)
library(caret)

library(devtools)
#install_github("vqv/ggbiplot")
library(ggbiplot)

pcacan <- prcomp(Cannabis, scale = TRUE)
print(pcacan)
summary(pcacan)
pcacan$center
dim(pcacan$x)
```

Cannabis biplots
```{r}
#install.packages("FactoMineR")
library("FactoMineR")
res.pca.can <- PCA(Cannabis, scale= TRUE)

#install("factoextra")
library("factoextra")
factoextra::fviz_pca_biplot(res.pca.can, axes = c(1, 2), col.ind = "black", col.var = "steelblue",repel = TRUE, select.var = list(contrib = 5),  title = "Cannabis PCA - Biplot")

```
For the Cannabis consumption data, we end up with 16 different PC models. We can see that most of the variation is expressed by PC1, PC2, and PC3. Looking closely at these models, it is noticeable that the CAN_20 variables (ranging from A to I) are the most important contributors to the first PC. The second PC is also affected by a range of variables, from CAN_25AGH to CAN_25E. 
The third PC is best explained by the CAN_05 variable with some influence from the CAN_05 and CAN_10 variables. It is also notable that 95% of the total variation in the dataset is expressed by the three first PC’s.  
The biplot for the Cannabis PCA allows us to see the displayed variants explained by the PC’s. We can see that the first component explains 62% of the variability in the dataset, whereas the second component explains 26% of the variability. 

```{r}
#compute standard deviation of each principal component
pcacan_std_dev <- pcacan$sdev

#compute variance
pcacan_var <- pcacan_std_dev^2

#proportion of variance explained
pcacan_prop_varia<- pcacan_var/sum(pcacan_var)
pcacan_prop_varia[1:17]

#scree plot
plot(pcacan_prop_varia, xlab = "Principal Component Cannabis",
             ylab = "Proportion of Variance Explained",
             type = "b")

#cumulative scree plot
plot(cumsum(pcacan_prop_varia), xlab = "Principal Component Cannabis",
              ylab = "Cumulative Proportion of Variance Explained",
              type = "b")

```
In order to find which components explain the highest variance, we computed the proportion of variance explained by dividing the variance by sum of the total variance. The output shows us that the first (62%) and second (26%) components explain the highest variance. When illustrated in a scree plot, we can see that we have reduced the predictors for Cannabis consumption from 16 to 5, without compromising the explained variance. Additionally, our cumulative variance plot shows that the 6 selected components produce a variance of about 99%. Our next steps would be to select these components and produce a predictive model.


Alcohol PCA
```{r}
pcaalc <- prcomp(Alcohol, scale=TRUE)
summary(pcaalc)
print(pcaalc)
pcaalc$center
dim(pcaalc$x)
```

Alcohol biplots
```{r}
res.pca.alc <- PCA(Alcohol, scale=TRUE)

factoextra::fviz_pca_biplot(res.pca.alc,  axes = c(1, 2), col.ind = "black", col.var = "steelblue",repel = TRUE, select.var = list(contrib = 5),  title = "ALcohol PCA - Biplot")
```

The Alcohol data produces 17 PC models. The most variation in Alcohol consumption is also expressed by PC1, PC2, PC3, containing 98% of the variation in the dataset. We can see that the most important contributors to the first PC are the ALC_25 variables (ranging from A to G). The second PC is affected by a different range of variables, those belonging to the ALC_30 group (ranging from A to F). And, the third PC is almost equally affected by the variables ALC_05, ALC_10C, and ALC_15C. 
The variants displayed in the Alcohol PCA biplot show us that the first component explains 48% of the variability in the dataset, whereas the second component explains 36% of the variability. 

```{r}
#compute standard deviation of each principal component
pcaalc_std_dev <- pcaalc$sdev

#compute variance
pcaalc_var <- pcaalc_std_dev^2

#proportion of variance explained
pcaalc_prop_varia<- pcaalc_var/sum(pcaalc_var)
pcaalc_prop_varia[1:17]

#scree plot
plot(pcaalc_prop_varia, xlab = "Principal Component Alcohol",
             ylab = "Proportion of Variance Explained",
             type = "b")

#cumulative scree plot
plot(cumsum(pcaalc_prop_varia), xlab = "Principal Component Alcohol",
              ylab = "Cumulative Proportion of Variance Explained",
              type = "b")
```
By computing the proportion of variance explained, we can see that the components explaining the highest variance are the first (48 %) and the second (36%). Our scree plot shows that by using PCA we have reduced the predictors in the data set from 17 to 4 without compromising our explained variance. We can see in the cumulative variance plot that our 4 selected components result in a variance close to 99%. Our next steps would be to select these components and produce a predictive model. 

 
###Conclusion###

